\documentclass[11pt]{article}

\usepackage{fontspec}
\usepackage[margin=1in]{geometry}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{amsmath,amssymb}
\usepackage{enumitem}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{tabularx}

% ── Listing styles ────────────────────────────────────────────────────
\definecolor{leanKeyword}{RGB}{0,0,180}
\definecolor{leanComment}{RGB}{100,100,100}
\definecolor{leanString}{RGB}{0,120,0}
\definecolor{leanOutput}{RGB}{60,60,60}
\definecolor{codebg}{RGB}{248,248,248}

\lstdefinelanguage{Lean}{
  morekeywords={import,open,namespace,end,def,theorem,example,syntax,
    let,match,return,try,catch,do,if,then,else,fun,where,with,
    partial,private,structure,deriving,in,for,mut,section,
    attribute,initialize,class,instance},
  morekeywords=[2]{use,resolve},
  keywordstyle=\color{leanKeyword}\bfseries,
  keywordstyle=[2]\color{leanKeyword}\bfseries,
  sensitive=true,
  morecomment=[l]{--},
  morecomment=[n]{/-}{-/},
  commentstyle=\color{leanComment}\itshape,
  morestring=[b]",
  stringstyle=\color{leanString},
  basicstyle=\ttfamily\small,
  breaklines=true,
  showstringspaces=false,
}

\lstset{
  language=Lean,
  backgroundcolor=\color{codebg},
  frame=single,
  rulecolor=\color{black!20},
  xleftmargin=2em,
  framexleftmargin=1.5em,
  numbers=left,
  numberstyle=\tiny\color{black!40},
  numbersep=8pt,
  aboveskip=1em,
  belowskip=1em,
}

\lstdefinestyle{output}{
  language={},
  basicstyle=\ttfamily\small\color{leanOutput},
  frame=single,
  rulecolor=\color{black!15},
  backgroundcolor=\color{white},
  numbers=none,
  xleftmargin=2em,
  framexleftmargin=0em,
}

\lstdefinestyle{pseudo}{
  language={},
  basicstyle=\ttfamily\small,
  frame=single,
  rulecolor=\color{black!20},
  backgroundcolor=\color{codebg},
  numbers=none,
  xleftmargin=2em,
  framexleftmargin=0em,
}

\lstdefinestyle{json}{
  language={},
  basicstyle=\ttfamily\small,
  frame=single,
  rulecolor=\color{black!20},
  backgroundcolor=\color{codebg},
  numbers=none,
  xleftmargin=2em,
  framexleftmargin=0em,
}

% ── Metadata ──────────────────────────────────────────────────────────
\title{\textbf{Content-Addressing Formal Mathematics}\\[4pt]
\large Stable, Decentralized Identity for Verified or Open Results}
\author{Marcello Paris\\\small\texttt{marcello.paris@gmail.com}}
\date{March 2026}

\begin{document}

\maketitle

\begin{abstract}
We propose \emph{content-addressing} for formal mathematics:
producing a stable identifier determined by mathematical content rather
than by human-chosen names, by computing a cryptographic hash of a declaration's canonicalized type expression.

Globally addressing mathematical facts enable a new paradigm for decentralized knowledge sharing, contribution and discovery.
We build a decentralized registry on top of these addresses, where the Lean verifier itself serves as the consensus mechanism.

We present a working implementation for Lean~4, evaluate it on some Mathlib standard modules,
and discuss the design space, limitations, and connections to
content-addressing in software engineering and to prior work on
cross-prover interoperability.
\end{abstract}

\tableofcontents
\bigskip


% =====================================================================
\section{Introduction}
\label{sec:intro}
% =====================================================================

Software engineers are familiar with content-addressing.  Git identifies
every commit, tree, and blob by the SHA-1 hash of its
content~\cite{git}.  The Nix package manager derives store paths from
the hash of all build inputs~\cite{dolstra2004nix}.  IPFS addresses
files by their cryptographic hash, making retrieval
location-independent~\cite{benet2014ipfs}.  The Unison programming
language names functions by the hash of their syntax
tree~\cite{unison}.

In each case, the core idea is the same: derive the identifier from the
\emph{content itself}, not from a human-chosen name or a filesystem
location.  The result is an identifier that is \emph{stable} (it does
not change when the content moves), \emph{universal} (it requires no
central authority), and \emph{self-authenticating} (anyone can verify it
by rehashing).

Formal mathematics has not yet adopted this pattern.  Theorems in proof
assistants---Lean~\cite{lean4}, Coq~\cite{coq}, Isabelle~\cite{isabelle},
Agda~\cite{agda}---are identified by fully qualified names tied to
project structure, file paths, and namespace conventions.  Renaming a
file, moving a declaration to a different module, or reorganizing a
namespace hierarchy breaks every downstream reference, even though the
mathematical content is unchanged.  This problem is not hypothetical: it
occurs routinely in Mathlib, the largest Lean~4 library, which undergoes
continuous refactoring across its ${\sim}200{,}000$
declarations~\cite{mathlib4}.  A complementary line of work addresses
the downstream consequence---automated proof
repair~\cite{ringer2021repair, gopinathan2023repair}---but does not
eliminate the root cause: names are not intrinsic to the mathematics.

This paper asks: \emph{can we apply content-addressing to formal
mathematics?}  We show that the answer is yes, with important caveats
about the granularity of equivalence (Section~\ref{sec:equiv}).  We
describe a working implementation for Lean~4 and a decentralized
registry architecture that requires no infrastructure beyond Git.

\paragraph{Contributions.}

\begin{itemize}[nosep]
  \item A content-addressing pipeline for Lean~4 declarations:
    canonicalization, serialization, and SHA-256 hashing
    (Section~\ref{sec:pipeline}).
  \item Two canonicalization levels with different
    stability/equivalence trade-offs (Section~\ref{sec:canon}).
  \item A Merkle DAG hashing mode that gives declarations truly
    content-determined identity across project boundaries
    (Section~\ref{sec:modes}).
  \item A decentralized registry architecture exploiting the
    monotonicity of mathematical proof (Section~\ref{sec:registry}).
  \item Empirical evaluation on 57{,}613 theorems from a Mathlib
    module environment (Section~\ref{sec:results}).
  \item An open-source implementation as a Lean~4 library and CLI
    tool (Section~\ref{sec:impl}).
\end{itemize}

% --------------------------------------------------------------------
\section{Name-based identity vs content-addressing}
% --------------------------------------------------------------------

% =====================================================================
\paragraph{Name-based Identity}
% =====================================================================

To use a theorem in Lean~4, a programmer writes something like:

\begin{lstlisting}[caption={Using a theorem by its fully qualified name}]
import Mathlib.Data.Finset.Sum

example : forall (s : Finset Nat) (f g : Nat -> Int),
    s.sum (f + g) = s.sum f + s.sum g :=
  Finset.sum_add_distrib
\end{lstlisting}

This single reference, \texttt{Finset.sum\_add\_distrib}, encodes three
pieces of information beyond the mathematics: (1)~the \textbf{project}
(Mathlib), (2)~the \textbf{file}
(\texttt{Mathlib/Data/Finset/Sum.lean}, needed for the \texttt{import}),
and (3)~the \textbf{namespace path}
(\texttt{Finset.sum\_add\_distrib}, a human-chosen convention).
None of these are intrinsic to the mathematical statement.  The theorem
says that finite sums distribute over addition---a fact that is true
independently of what we call it or where we store it.

When any of these coordinates change, references break.  A theorem
renamed for consistency (\texttt{sum\_add\_distrib} $\to$
\texttt{sum\_add}), a declaration relocated to a different module
(\texttt{Data.Finset.Sum} $\to$ \texttt{Algebra.BigOps}), a namespace
hierarchy flattened or reorganized, a project forked with new
prefixes---each is a no-op mathematically, but a breaking change
syntactically.  The cost is borne by every downstream project: CI
pipelines fail, students' homework stops compiling, automated tooling
must track name migrations.

Within a single library like Mathlib, naming conventions and import
paths are at least centrally managed.  Across independent projects, the
situation is worse.  There is no standardized way for project~A to
refer to a result proved in project~B, other than adding a direct
dependency on B's entire package---including its specific version, file
layout, and naming scheme.  If two independent projects prove the same
theorem under different names, there is no mechanism to recognize that
they state the same mathematical fact.  Each project is an island, and
the only bridges are explicit package dependencies with name-level
coupling.

This problem will only grow.  Mathlib is evolving into the shared
foundation that virtually all Lean projects depend on, but the
ecosystem is expanding well beyond pure mathematics: formalization
projects in cryptography, blockchain protocol verification, physics,
and other applied domains are emerging, each building on Mathlib and
on each other.  As the number of independent projects grows, so does
the cost of name-level coupling---every cross-project reference is a
fragile link that any upstream refactoring can break.


% --------------------------------------------------------------------
\paragraph{Content-Addressing}
% --------------------------------------------------------------------

The central idea is simple: \emph{hash the mathematics, not the name.}

Given a declaration with a type expression (its mathematical statement),
we: (1)~\textbf{canonicalize} the type expression to remove
representation artifacts, (2)~\textbf{serialize} the canonical
expression to a byte sequence, and (3)~\textbf{hash} the byte sequence
with SHA-256.  The resulting 256-bit hash is the declaration's
\emph{content address}.
For theorems, the content address is therefore determined entirely by
the statement---the proof term is never hashed.
For propositions, this is inevitable given proof irrelevance.

Content addresses have four key properties.  They are \emph{stable}:
the address does not change when a declaration is renamed, moved, or
reorganized.  They are \emph{universal}: no project, file, or import
context is needed, and no coordination or central authority is required.
They are \emph{self-authenticating}: anyone can verify an address by
recomputing the hash from the declaration's type.  And they are
\emph{collision-resistant}: two mathematically different statements will
(with overwhelming probability) have different addresses.

An important caveat: content addresses identify declarations that share
the same \emph{canonical type}, but this does not make them freely
interchangeable.  Declarations carry metadata (attributes, instance
registrations) that the hash does not capture, and definitions with the
same type may compute differently.  We discuss these limits in
Section~\ref{sec:equiv}.

% =====================================================================
\section{A working pipeline}
\label{sec:pipeline}
% =====================================================================

The address of a declaration is computed in three stages:
canonicalization, serialization, and hashing.

% --------------------------------------------------------------------
\subsection{Canonicalization}
\label{sec:canon}
% --------------------------------------------------------------------

Canonicalization transforms a Lean~4 \texttt{Expr} (expression) into a
canonical form where expressions differing only in irrelevant ways
become syntactically identical.  We implement two levels, offering a
trade-off between stability and the strength of the equivalence
relation.

\paragraph{Level~0: structural canonicalization.}
Level~0 is a pure transformation (no monadic context required) that
normalizes two sources of superficial variation.  First, \emph{universe
parameters}: Lean uses named universe polymorphism, and these names are
arbitrary; Level~0 renames them to positional indices (\texttt{u\_0},
\texttt{u\_1}, \ldots) by order of first occurrence.  Second,
\emph{metadata annotations}: Lean attaches \texttt{.mdata} nodes to
expressions for internal bookkeeping (source positions, elaboration
hints); these carry no mathematical content and are stripped.

Lean~4 uses de~Bruijn indices~\cite{debruijn1972} for bound variables,
so $\alpha$-equivalence is handled by the representation itself.
Binder \emph{names} (e.g., \texttt{n} vs.\ \texttt{x}) are cosmetic
and excluded from hashing.  Level~0 is fast and deterministic.  It
depends only on the structure of the \texttt{Expr} type, which is
stable within a major Lean version but has changed across major versions
(e.g., Lean~3 to Lean~4).  Its limitation is that it does not normalize
definitional equalities: two formulations that differ by an unfolded
abbreviation or a reduced beta-redex receive different hashes.

\paragraph{Level~1: reducible normalization.}
Level~1 extends Level~0 with \emph{reducible-transparency
normalization}.  It requires a \texttt{MetaM} (meta-programming monad)
context because it invokes Lean's \texttt{whnf} (weak head normal form)
reduction.  The normalizer, \texttt{deepNormalize}, recursively applies
\texttt{whnf} at every subexpression under \texttt{reducible}
transparency:

\begin{lstlisting}[style=pseudo,caption={Deep normalization (simplified pseudocode)}]
deepNormalize(e, depth):
  if depth >= 256: return e       -- recursion guard
  e <- whnf e                     -- reducible transparency
  match e:
    .app f a      -> .app (deepNormalize f) (deepNormalize a)
    .lam _ t b _  -> normalize t; open binder; normalize body;
                     re-abstract
    .forallE _ t b _ -> same as lambda
    .letE _ _ v b -> normalize v; substitute; normalize result
    .mdata _ e    -> deepNormalize e
    .proj _ _ e   -> normalize e; whnf the projection
    _             -> return e
\end{lstlisting}

Under reducible transparency, notation aliases (e.g.,
\texttt{HAdd.hAdd} $\to$ \texttt{Add.add}), declarations marked
\texttt{@[reducible]}, beta-redexes, let-bindings, and reducible
structure projections are all unfolded.  Crucially,
\texttt{@[semireducible]} declarations (including typeclass instances)
are \emph{not} unfolded, preserving abstraction boundaries:
\texttt{Add Nat} stays as \texttt{Add Nat} rather than being replaced
by its instance implementation.

As an example, consider \texttt{theorem add\_comm\_v1 : $\forall$ (a b
: Nat), a + b = b + a} and a variant \texttt{add\_comm\_v2 :
$\forall$ (a b : Nat), b + a = a + b}.  At Level~0, these may receive
different hashes if the elaborated expressions differ structurally.  At
Level~1, after \texttt{whnf} normalization, both reduce to the same
canonical form and receive the same hash.  On
\texttt{Init.Data.Nat.Basic}, the largest equivalence class grows from
391 members (Level~0) to 467 (Level~1)---76~additional declarations
collapsed by reducible unfolding.  Section~\ref{sec:results} presents a
larger-scale evaluation.

\paragraph{Level~2: full definitional equality (open problem).}
The ideal canonicalization would assign the same hash to all
definitionally equal types.  However, Lean's dependent type theory does
not admit unique normal forms under full definitional
equality~\cite{abel2018decidability}.  Achieving Level~2 would require
either restricting the fragment of the type theory considered, or
developing new normalization techniques.  We leave this as an open
problem (see Section~\ref{sec:equiv} for further discussion).

% --------------------------------------------------------------------
\subsection{Serialization, Hashing, and Hashing Modes}
\label{sec:serial}
% --------------------------------------------------------------------

\paragraph{Serialization.}
The canonical expression is serialized to a \texttt{ByteArray} using a
tagged encoding: each \texttt{Expr} constructor is assigned a unique tag
byte, followed by the serialization of its children.  Binder names are
omitted (de~Bruijn indices suffice).  \texttt{BinderInfo} (implicit,
explicit, instance, etc.)\ is included for lambda and forall nodes.
\texttt{.mdata} is transparent: the child is serialized directly.  For
\texttt{.const name levels}, the treatment of \texttt{name} depends on
the hashing mode (below).

\paragraph{Hashing.}
The serialized byte array is hashed with SHA-256, producing a
64-character hexadecimal string.  We use SHA-256 via OpenSSL's EVP API
through a minimal C FFI shim, chosen for ubiquity and ease of binding
rather than performance (a faster hash such as BLAKE3 would work equally
well---the choice of hash function is orthogonal to the addressing
design).

\paragraph{Hashing modes.}
\label{sec:modes}
When serializing a constant reference (\texttt{.const name levels}),
there is a choice of what to emit for \texttt{name}.  In
\emph{name-based} mode (default), the declaration's \texttt{Name} is
serialized as a UTF-8 string---fast and stable within a single library,
but if a dependency is renamed, the address changes.  In
\emph{content-based} (Merkle DAG) mode, the 32-byte content hash of the
referenced declaration is emitted instead.  Declarations are processed
in topological (dependency) order via Kahn's
algorithm~\cite{kahn1962topological}, so each hash incorporates the
hashes of its transitive dependencies.  The result is a Merkle
DAG~\cite{merkle1988} where a declaration's identity depends only on the
mathematical content of everything it (transitively) references.

Mutually recursive declarations form cycles in the dependency graph.
The topological sort detects these and falls back to name-based hashing
for cycle members---a conservative choice that preserves soundness at
the cost of weaker identity for mutual definitions.  Mutual recursion
is uncommon in theorem \emph{statements}, but Lean's elaboration of
inductive types generates mutually recursive auxiliary declarations
(\texttt{rec}, \texttt{casesOn}, etc.), so the fallback affects more
declarations than one might initially expect when indexing definitions
alongside theorems.

\begin{table}[h]
\centering
\begin{tabular}{@{}lll@{}}
  \toprule
  \textbf{Mode} & \textbf{Constant references become} & \textbf{Trade-off} \\
  \midrule
  Name-based & The declaration's \texttt{Name} string & Fast; changes if deps renamed \\
  Content-based & 32-byte content hash of the dep & True content identity; slower \\
  \bottomrule
\end{tabular}
\caption{Hashing modes.}
\label{tab:modes}
\end{table}


% --------------------------------------------------------------------
\subsection{Results}
\label{sec:results}
% --------------------------------------------------------------------

This section presents both concrete examples of content-addressing in
action and a quantitative evaluation on Mathlib.

\paragraph{Duplicate detection.}
Large libraries inevitably accumulate duplicates---theorems with
different names but identical mathematical content.  Content-addressing
detects these automatically: duplicates share the same hash.  The
technique of comparing type expressions for deduplication is already
used in the AI-for-theorem-proving community---for instance, synthetic
theorem generation pipelines deduplicate candidates by \texttt{Expr}
equality~\cite{syntheticthm}.  Content-addressing provides a
principled, persistent, and cross-project version of what is currently
done ad~hoc within individual pipelines.

\paragraph{Cross-project references.}
Consider two independent projects that both work with a mathematical
concept:

\begin{lstlisting}[caption={Two projects proving the same lemma independently}]
-- In project-a/Algebra/Basic.lean
theorem add_comm_nat : forall a b : Nat, a + b = b + a :=
  Nat.add_comm

-- In project-b/MyMath/Nat.lean
theorem nat_addition_commutative : forall a b : Nat, a + b = b + a :=
  fun a b => Nat.add_comm a b
\end{lstlisting}

Despite different names, files, and projects, both declarations receive
the same content address.  A search tool can discover that these are the
same mathematical fact, enabling cross-project deduplication and
discovery without any coordination between the projects.

\paragraph{Open problems and conditional results.}
Content-addressing supports a structured workflow for open problems,
though this aspect of the system is prospective---the mechanism is
implemented but has not yet been exercised at scale.  A researcher
publishes a conjecture as an \emph{open point}---a \texttt{Prop} with a
content address but no proof:

\begin{lstlisting}[caption={An open conjecture and a conditional result}]
@[open_point "Bertrand's postulate, elementary form"]
def BertrandPostulate : Prop :=
  forall n : Nat, n >= 1 -> exists p, Nat.Prime p /\ n < p /\ p < 2 * n

@[publish "Consequence of Bertrand's postulate"]
theorem prime_between (n : Nat) (h : BertrandPostulate) (hn : n >= 1) :
    exists p, Nat.Prime p /\ n < p /\ p < 2 * n :=
  h n hn
\end{lstlisting}

The registry (Section~\ref{sec:registry}) tracks which conditional
results depend on which open points.  When someone eventually proves the
conjecture, the status of all downstream results cascades from
``conditional'' to ``proved.''  Several practical questions remain open:
how to handle conjectures that turn out to be false (or independent),
how to distinguish \texttt{sorry}-based development from genuinely
axiomatic assumptions, and how to curate status metadata across
projects.  We discuss these further in Section~\ref{sec:future}.

\paragraph{Registry generation.}
A project publishes its annotated declarations by adding
\texttt{\#ca\_registry "registry/"} at the end of a Lean file that
imports the annotated modules.  When \texttt{lake build} compiles this
file, it writes \texttt{declarations.json} (address, name, status,
type, deps for each declaration) and \texttt{meta.json} (project
summary with counts by status) to the specified directory.

\paragraph{Evaluation on Mathlib.}
We evaluated the pipeline on the environment loaded by
\texttt{Mathlib.Algebra.Group.Basic}, a core module of Mathlib that
transitively imports 1{,}209 modules covering foundational arithmetic,
data structures, and algebraic hierarchy definitions.  The pipeline was
run in name-based mode at Level~1 (reducible normalization), restricted
to theorems.  All 57{,}613 theorems in the transitive environment were
processed successfully---no declarations were skipped or failed.

The 57{,}613 theorems produce 57{,}036 unique type addresses.  Of
these, 544 addresses are shared by more than one theorem, covering
1{,}121 theorems (1.9\% of the total).

\begin{table}[h]
\centering
\begin{tabular}{@{}rrr@{}}
  \toprule
  \textbf{Class size} & \textbf{Classes} & \textbf{Theorems} \\
  \midrule
  1 (unique)  & 56{,}492 & 56{,}492 \\
  2           & 513      & 1{,}026 \\
  3           & 29       & 87 \\
  4           & 2        & 8 \\
  \midrule
  Total       & 57{,}036 & 57{,}613 \\
  \bottomrule
\end{tabular}
\caption{Equivalence class size distribution (L1, name-based, theorems only).}
\label{tab:eval}
\end{table}

\paragraph{Examples of detected duplicates.}
The equivalence classes reveal genuine duplicates---theorems that exist
under multiple names due to historical renaming, namespace aliasing, or
independent re-statement:

\begin{lstlisting}[style=output,caption={Selected equivalence classes (L1)}]
Size 4:
  by_contra = by_contradiction
           = Classical.byContradiction = Classical.by_contradiction
  Type: (Not p -> False) -> p

Size 3:
  congr_arg = congrArg = let_val_congr
  Type: forall (b : a -> b), a = a' -> b a = b a'

  Nat.add_sub_cancel = Nat.add_sub_cancel_right
                     = Nat.add_sub_self_right
  Type: forall (a b : Nat), a + b - b = a

Size 2:
  mul_one = MulOneClass.mul_one
  Type: forall [MulOneClass M] (a : M), a * 1 = a

  mul_comm = CommMagma.mul_comm
  Type: forall [CommMagma G] (a b : G), a * b = b * a
\end{lstlisting}

The types shown are simplified pretty-printed representations of the
canonical forms; binder names and implicit arguments are artifacts of
the printer, not of the hash.
These duplicates arise from distinct sources.  The \texttt{by\_contra}
cluster contains four names for the same classical reasoning principle,
accumulated as Lean's standard library evolved.  The \texttt{mul\_one} /
\texttt{MulOneClass.mul\_one} pair reflects the algebraic hierarchy: the
root name and the typeclass-qualified name coexist as aliases.  The
\texttt{Nat.add\_sub\_cancel} triple shows the same arithmetic fact
named by three conventions.  In all cases, the content address correctly
identifies these as stating the same mathematical fact, modulo Level~1
canonicalization.

\paragraph{Coverage.}
All 57{,}613 theorems were hashed without error, demonstrating that
the pipeline (including the micro-batching workaround for
\texttt{whnf}; Appendix~\ref{app:engineering}) handles the full
transitive closure of a Mathlib module.  The current evaluation covers
a single module's environment at Level~1 only.  A systematic comparison
of Level~0 vs.\ Level~1 across multiple Mathlib entry points, and
evaluation in content-based (Merkle DAG) mode, remain as future work.



% =====================================================================
\section{Equivalence Classes and Their Limits}
\label{sec:equiv}
% =====================================================================

Two declarations sharing the same content address state the same
mathematical fact (modulo the canonicalization level).  The set of all
declarations with a given address forms an \emph{equivalence class}.

\paragraph{Same hash does not mean interchangeable.}  The content
address captures the \emph{type}, not the proof term, runtime behavior,
or elaboration metadata.  This has important consequences.
\emph{Theorems in direct proofs} are interchangeable, by proof
irrelevance.  \emph{Theorems in tactics} are not: a tactic like
\texttt{simp [Nat.add\_mul]} relies on the \texttt{@[simp]} attribute
attached to that specific \emph{name}; substituting an alias with the
same hash but different attributes would change the tactic's behavior.
\emph{Definitions} are not interchangeable: two definitions with the
same type may compute differently.  \emph{Instances} are not
interchangeable: typeclass instances are registered by name and
participate in instance resolution.

For this reason, resolution is always \emph{by name}.  The content
address is a stable, project-independent key for \emph{discovery} and
\emph{cross-referencing}, but resolution goes through a specific
declaration name that carries the necessary metadata.

\begin{lstlisting}[style=output,caption={Alias discovery via content address}]
Resolved: Nat.add_mul
  Hash:   e1ea8f392f15...  (L1)
  Module: Init.Data.Nat.Basic
  Kind:   theorem
  Type:   forall (n m k : Nat), (n + m) * k = n * k + m * k
  Aliases (1): Nat.right_distrib
\end{lstlisting}

\paragraph{Canonicalization and definitional equality.}
The fundamental question underlying the equivalence relation is: which
expressions should receive the same hash?  The ideal answer is ``all
definitionally equal expressions,'' but this is not achievable in
general.  Lean's Calculus of Inductive Constructions (CIC) with proof
irrelevance does not admit unique normal forms under full definitional
equality~\cite{abel2018decidability, coquand1991algorithm}.  The
reduction rules---$\beta$, $\delta$ (at various transparencies),
$\iota$, $\zeta$, and proof irrelevance---do not confluently normalize
all expressions to a unique representative.

Our two-level approach is a pragmatic compromise.  Level~0 captures
syntactic equality modulo universe renaming and metadata---a
conservative but stable equivalence.  Level~1 captures equality modulo
reducible $\delta$-reduction, $\beta$-reduction, $\zeta$-reduction, and
reducible $\iota$-reduction---a strictly coarser equivalence that
collapses common aliases.  The gap between Level~1 and full definitional
equality is large: semireducible and opaque definitions are not
unfolded, so type synonyms at these transparencies create distinct
hashes.  In principle, one could add intermediate levels (semireducible
transparency, instance unfolding), but each level increases computation
cost, reduces stability across Lean versions, and risks divergent
normalization.

\paragraph{Universe polymorphism.}
Lean uses \emph{named} universe polymorphism: declarations are
parameterized over universe level variables with programmer-chosen
names.  Two declarations with identical bodies but different universe
parameter names (e.g., \texttt{.\{u, v\}} vs.\ \texttt{.\{a, b\}})
should hash identically.  Level~0 handles this by renaming universe
parameters to positional indices: the first universe variable
encountered becomes \texttt{u\_0}, the second \texttt{u\_1}, and so on.
The traversal order is deterministic (left-to-right depth-first over
the type expression).


% =====================================================================
\section{Implementation}
\label{sec:impl}
% =====================================================================

The implementation is a Lean~4 package (\texttt{CA}) available at
\url{https://github.com/marcellop71/CA} (repository to be made public
upon acceptance).  Downstream projects depend on it with a single line:

\begin{lstlisting}[style=pseudo]
require ca from git "https://github.com/marcellop71/CA" @ "main"
\end{lstlisting}

\paragraph{End-to-end workflow.}  A user who wants to content-address
their project performs four steps: (1)~add \texttt{require ca} to their
\texttt{lakefile.lean} and run \texttt{lake update}; (2)~annotate
declarations with \texttt{@[publish]} or \texttt{@[open\_point]}
(directly or retroactively via \texttt{attribute [publish] SomeName});
(3)~add \texttt{\#ca\_registry "registry/"} to a Lean file that
imports the annotated modules; (4)~run \texttt{lake build}---the
registry folder is generated automatically during compilation, no
separate tool invocation required.  The generated \texttt{registry/}
folder is committed to Git and pushed; the project is now a node in the
decentralized registry.

\paragraph{CLI.}  For batch processing (e.g., indexing all of Mathlib),
a CLI executable provides \texttt{ca address} (compute hashes, export
JSON manifest and TSV edge list) and \texttt{ca fetch} (store to
Redis).  Redis serves as an interactive lookup layer: given a content
address, it returns all known names, types, and modules in
constant time---a capability the static JSON manifest does not provide
efficiently at scale.

\begin{table}[h]
\centering
\begin{tabular}{@{}ll@{}}
  \toprule
  \textbf{Module} & \textbf{Role} \\
  \midrule
  \texttt{CA.Canonical}           & L0 (pure) and L1 (MetaM) canonicalization \\
  \texttt{CA.SHA256}              & SHA-256 via OpenSSL EVP (C FFI) \\
  \texttt{CA.ExprHash}            & Expr serialization, batch hashing, Merkle DAG \\
  \texttt{CA.Export}              & JSON manifest, TSV edge list, statistics \\
  \texttt{CA.Util}                & Dependency collection, declaration classification \\
  \texttt{CA.Registry.Basic}      & Entry types, persistent extensions, query API \\
  \texttt{CA.Registry.Attributes} & \texttt{@[publish]}, \texttt{@[open\_point]} registration \\
  \texttt{CA.Registry.Generate}   & \texttt{\#ca\_registry} command, JSON output \\
  \bottomrule
\end{tabular}
\caption{Module overview.}
\label{tab:modules}
\end{table}

\paragraph{Build and dependencies.}  The default \texttt{lake build}
compiles the \texttt{CA} library, which requires only OpenSSL.  The CLI
executable (\texttt{lake build ca}) additionally requires hiredis for
Redis-backed lookups.  A Nix flake provides a development shell with the
Lean toolchain and all native dependencies for both Linux and macOS.
The implementation currently targets Lean~4.29.0-rc1 and tracks the
Lean release cycle; it depends on the Batteries
library~\cite{batteries} for \texttt{NameMapExtension}.


% =====================================================================
\section{Limitations and Future Work}
\label{sec:limits}
% =====================================================================

\paragraph{Current limitations.}
Level~0 and Level~1 canonicalization do not capture full definitional
equality; the gap between Level~1 and the ideal is large and may be
fundamentally unclosable.  Level~1 addresses depend on Lean's
\texttt{whnf} behavior, which can change across toolchain versions;
Level~0 addresses are stable within a major Lean version, while Level~1
should be considered best-effort within a given toolchain.  Lean's
native \texttt{whnf} (implemented in C) lacks heartbeat checks and
crashes on large environments; a micro-batching workaround is effective
but inelegant (Appendix~\ref{app:engineering}).  Applying \texttt{whnf}
to definition values causes process-level memory growth that crashes
regardless of batch boundaries, limiting content-addressing of
definitions (as opposed to theorems, where proof irrelevance makes value
hashing unnecessary).  Finally, Lean~4's elaboration pipeline does not
support user-defined fallbacks for unresolved names; our resolution
mechanism requires explicit syntax (\texttt{resolve\%}), preventing
seamless integration.

\paragraph{Future work.}
\phantomsection\label{sec:future}
Several directions are worth pursuing.  \emph{Resolution elaborators}:
implement \texttt{use}, \texttt{use!}, and \texttt{resolve\%} as Lean~4
elaborators that query the registry, resolve by content address, and
dynamically import modules.  The intended syntax (not yet implemented):

\begin{lstlisting}[caption={Prospective resolution syntax (not yet implemented)}]
-- resolve% looks up a content address in the registry and
-- returns the name of a matching declaration in scope
example : forall a b : Nat, a + b = b + a :=
  resolve% "a1b2c3d4e5f6..."   -- resolves to Nat.add_comm
\end{lstlisting}

\emph{Ecosystem-wide indexing}: apply
content-addressing to the entirety of Mathlib to produce a global index
of equivalence classes, enabling large-scale duplicate detection,
library health metrics, and structured navigation.
\emph{Open-problem workflows}: develop the open-point and
conditional-result mechanism (Section~\ref{sec:results}) into a robust
workflow, addressing false conjectures, \texttt{sorry}-based vs.\
axiomatic assumptions, and cross-project status curation.
\emph{AI-assisted proof search}: content addresses provide a stable,
type-based index for machine learning models to retrieve relevant lemmas
by querying the content address of a goal type.
\emph{Cross-system alignment}: explore content-addressing in other proof
assistants (Coq, Isabelle, Agda) and investigate whether addresses can
be made comparable via shared canonicalization.
\emph{Intermediate canonicalization levels}: investigate canonicalization
at semireducible transparency or with selective instance unfolding.
\emph{Formal verification}: prove (in Lean itself) that the
canonicalization and serialization stages preserve the intended
equivalence relation.

% =====================================================================
\section{Decentralized Registry}
\label{sec:registry}
% =====================================================================

Content addresses give every declaration a stable,
location-independent identity.  The registry organizes the mutable
metadata---who proved what, what is open, what depends on what---around
this immutable core.

Every formalization in Lean already lives in a Git repository.  The
decentralized registry exploits this directly: each participating
project hosts a \texttt{registry/} folder in its own repo, and the
global registry is the union of all such folders.  No additional
infrastructure is needed---no IPFS, no blockchain, no central server.

This simplicity is possible because \textbf{mathematical consensus is
trivial}.  Unlike distributed systems that must agree on ordering or
resolve write conflicts, a mathematical registry has a monotonic truth
condition: a proposition is proved if \emph{any} repository contains a
valid proof.  Proofs cannot be retracted---once proved, always proved.
Two repositories proving the same proposition under different names is
not a conflict; it is redundancy, and the content address shows they
state the same fact.  There is nothing to vote on, no quorum to reach.
This observation connects to the broader vision of a pluralistic formal
library where multiple communities contribute to a shared body of
verified knowledge~\cite{kohlhase2017qed}.

In concrete terms:

\begin{itemize}[nosep]
  \item \textbf{Publication} is \texttt{git push}---a project generates
    its \texttt{registry/} folder, commits it, and pushes.
  \item \textbf{Verification} is \texttt{lake build}---Lean's kernel
    type-checks the proofs; anyone can clone and verify independently.
  \item \textbf{Consensus} is trivial: one valid proof is enough.
  \item \textbf{Discovery} is a curated list of known repositories
    (\texttt{sources.json}).
\end{itemize}

\paragraph{Annotation attributes.}
Authors annotate their code with two attributes provided by the
\texttt{CA.Registry} module.
\texttt{@[open\_point]} marks a \texttt{def X : Prop} as an open
problem---a mathematical statement published without a proof.
Applying it to a theorem or a non-\texttt{Prop} definition is a
compile-time error.
\texttt{@[publish]} marks any declaration (theorem, definition, axiom)
for publication to the registry.
Both attributes accept an optional description string and support
retroactive application from a separate file, allowing annotation of
existing code or upstream dependencies without modifying the original
source:

\begin{lstlisting}[caption={Retroactive annotation of existing declarations}]
import CA
import MyProject.Theorems

attribute [open_point "My conjecture"] MyProject.SomeConjecture
attribute [publish "Key lemma"]        MyProject.some_lemma
\end{lstlisting}

\paragraph{Declaration status.}
Status is computed automatically from the Lean environment:

\begin{table}[h]
\centering
\begin{tabular}{@{}lll@{}}
  \toprule
  \textbf{Pattern} & \textbf{Status} & \textbf{Rule} \\
  \midrule
  \texttt{theorem foo : P := proof} & proved & Type-checks, no \texttt{sorry} \\
  \texttt{def P : Prop} with \texttt{@[open\_point]} & open & Explicitly marked \\
  \texttt{theorem bar (h : OpenP) : Q} & conditional & Proved modulo open hypotheses \\
  \bottomrule
\end{tabular}
\caption{Automatic status classification.}
\label{tab:status}
\end{table}

For conditional declarations, the registry records which open addresses
are required---the declarations whose proof would promote the entry
from conditional to proved.

\paragraph{Registry folder.}
The \texttt{registry/} folder is generated locally inside each project.
The \texttt{\#ca\_registry "registry/"} command (or the \texttt{ca
registry} CLI) loads the compiled environment, collects annotated
declarations, computes content addresses, classifies status, and writes
two files:

\begin{itemize}[nosep]
  \item \texttt{declarations.json}---the append-only manifest.  Each
    entry contains the content address, name, module, kind, status,
    pretty-printed type, and type dependencies.
  \item \texttt{meta.json}---project-level metadata (hash level,
    open/proved/conditional counts).
\end{itemize}

Entries are never removed.  Status updates (open $\to$ proved) are
reflected by updating the status field in place; Git history provides
the audit trail.

\paragraph{Status cascade.}
When a proof is submitted for an open point, conditional declarations
downstream may become unconditionally proved:

\begin{enumerate}[nosep]
  \item A previously open address acquires a proof---status becomes
    \texttt{proved}.
  \item All \texttt{conditional} entries whose open dependencies
    included the newly proved address are re-evaluated.
  \item If all of an entry's open dependencies are now proved, it is
    promoted to \texttt{proved}.
  \item This cascades transitively.
\end{enumerate}

Cross-project cascades emerge when a consumer merges multiple
registries: an open point in project~A may be proved in project~B,
promoting conditional entries in project~C, with no coordination
between A, B, and~C.  The dependency graph also makes \emph{impact}
computable: the impact of an open point is the number of conditional
declarations transitively downstream of it.

\paragraph{Sources and discovery.}
Because the registry is fully decentralized, a discovery mechanism is
needed so that resolution tools know where to look.  A curated
\texttt{sources.json} file in the CA repository enumerates known Git
repositories that host a \texttt{registry/} folder:

\begin{lstlisting}[style=json,caption={\texttt{sources.json} format}]
{ "registries": [
    { "url": "https://github.com/leanprover-community/mathlib4",
      "registry_path": "registry/" },
    { "url": "https://github.com/user/my-project",
      "registry_path": "registry/" }
  ]
}
\end{lstlisting}

Adding a project to the ecosystem requires five steps: (1)~add
\texttt{require ca} to the project's \texttt{lakefile.lean};
(2)~annotate declarations with \texttt{@[publish]} or
\texttt{@[open\_point]}; (3)~add \texttt{\#ca\_registry "registry/"}
to a Lean file that imports the annotated modules;
(4)~\texttt{lake build}, commit \texttt{registry/}, and push;
(5)~open a PR adding one entry to \texttt{sources.json}.
Step~5 is the only coordination point; everything else is local to the
project.  The static list is the right starting point---it can later be
replaced by automatic discovery (e.g., a GitHub topic tag or a DHT)
without changing the rest of the architecture.

\paragraph{Current maturity.}  The architecture described here is
implemented but early-stage.  Discovery relies on a manually curated
\texttt{sources.json}, there is no mechanism for revoking or correcting
erroneous entries, and metadata conflicts (e.g., two projects assigning
different status to the same address) are resolved by the consumer at
merge time.  The theoretical elegance of monotonic consensus is real,
but the tooling around it is minimal.


% =====================================================================
\section{Related Work}
\label{sec:related}
% =====================================================================

\paragraph{Content-addressing in software and programming languages.}
Content-addressed storage is well-established in software systems
(Git~\cite{git}, IPFS~\cite{benet2014ipfs},
Nix~\cite{dolstra2004nix}; see Section~\ref{sec:intro} for an
overview).
Unison~\cite{unison} is the closest analog in the programming language
domain.  Unison was \emph{designed from the ground up} around
content-addressing: the entire codebase is stored as a content-addressed
database, names are metadata, and refactoring is a metadata operation
that never invalidates hashes.  Unison hashes the full abstract syntax
tree (the implementation), not just the type, and uses 512-bit SHA3
hashes.  Our setting differs in three important ways.  First, we
\emph{retrofit} content-addressing onto an existing language (Lean~4)
that was not designed for it, which creates friction---the
\texttt{resolve\%} syntax, the inability to hook into name resolution,
and the non-interchangeability of same-hash declarations
(Section~\ref{sec:equiv}) are all consequences of this.  Second, we hash
only the \emph{type expression}, not the implementation, which is
justified by proof irrelevance for theorems but means definitions with
the same type are not distinguished.  Third, canonicalization in a
dependently typed setting is complicated by definitional equality and
universe polymorphism in ways that simply-typed languages avoid.

\paragraph{Cross-prover library management.}
Hurd's OpenTheory project~\cite{hurd2011opentheory} is the closest
prior system in the formal mathematics world.  OpenTheory built a
package management system for the HOL family of provers (HOL4, HOL
Light, ProofPower) where theories are represented as composable,
portable packages with explicit assumption and conclusion interfaces.
Theories can be shared across HOL implementations via \emph{theory
interpretations} that map names between systems.  OpenTheory does not
use cryptographic hashing for identity, but it directly tackles the same
core problem: making theorems referenceable independently of the prover
and naming scheme.  Our approach differs in mechanism
(content-hashing vs.\ theory interpretation) and scope (within Lean~4
vs.\ across HOL variants), but the motivations are closely aligned.
Mathlib~\cite{mathlib4} employs a centralized monorepo model with
coordinated naming conventions.  The Coq ecosystem uses opam
packages~\cite{coq} for library distribution.  The Isabelle Archive of
Formal Proofs~\cite{isabelle_afp} uses a curated submission model.
None of these use content-addressing for declaration identity.

\paragraph{Concept alignment across proof assistants.}
The problem of recognizing equivalent mathematical concepts across
different logical foundations has been studied by Gauthier, Kaliszyk,
and Urban~\cite{gauthier2019aligning}, who developed pattern-matching
and scoring algorithms to align concepts across HOL4, Coq, and Mizar
libraries, and by M\"uller et al.~\cite{muller2017alignment}, who
studied alignment-based translations across formal systems.  These
approaches solve a harder problem than ours: alignment across
\emph{different type theories}, where syntactic comparison is not
directly meaningful.  Our content-addressing operates within a single
type theory, where canonical forms can be compared syntactically.
Intra-system content-addressing is a natural prerequisite for cross-system
alignment: identifying equivalent results within a system is a necessary
first step before comparing across systems.

\paragraph{Mathematical knowledge management.}
The OMDoc/MMT framework~\cite{rabe2013scalable, kohlhase2006omdoc}
provides a content-oriented representation for mathematical knowledge,
including a module system with morphisms between theories.  The vision
of a pluralistic, decentralized formal library has been articulated by
Kohlhase and Rabe~\cite{kohlhase2017qed}.  OpenMath content
dictionaries~\cite{openmath} standardize representations of
mathematical objects.  Our approach is complementary: rather than
standardizing representations \emph{a priori}, we derive identifiers
from the representations that already exist in proof assistants.

\paragraph{Proof repair.}
The downstream consequence of name-based identity---proofs breaking
under library refactoring---has motivated work on automated proof
repair.  Ringer et al.~\cite{ringer2021repair} repair proofs across
type equivalences in Coq, and Gopinathan et
al.~\cite{gopinathan2023repair} automate repair for verified library
upgrades.  Content-addressing and proof repair are complementary
strategies: the former aims to prevent breakage by decoupling identity
from naming, while the latter mitigates breakage after the fact.

\paragraph{Type-based deduplication and normalization.}
The technique of comparing \texttt{Expr} representations of theorem
types for deduplication is used in AI-for-theorem-proving
pipelines~\cite{syntheticthm}, essentially performing Level~0 of our
pipeline without cryptographic persistence.  The question of normal
forms in dependent type theory has a long history: Abel and
Scherer~\cite{abel2012long} study normalization by evaluation for
dependently typed languages, and the decidability of definitional
equality in CIC variants has been analyzed by
Coquand~\cite{coquand1991algorithm} and
others~\cite{abel2018decidability}.  Our work sidesteps the question of
full normalization by offering pragmatic partial normalization levels.
The Dedukti logical framework~\cite{assaf2016dedukti} aims to serve as
an interchange format between proof assistants, encoding different type
theories in a common language.


% =====================================================================
\section{Conclusion}
\label{sec:conclusion}
% =====================================================================

We have presented a content-addressing system for formal mathematics,
implemented as a Lean~4 library.  By hashing canonicalized type
expressions, we assign declarations stable identifiers determined by
their mathematical content rather than their human-chosen names.

Content-addressing does not require full normalization to be useful.
Even partial canonicalization (Level~0) provides meaningful stability
and cross-project discoverability.  The decentralized registry
architecture, made possible by the monotonicity of mathematical proof,
requires no infrastructure beyond Git---though significant engineering
work remains to make it practical at ecosystem scale.

The system is a working prototype, and important challenges
remain---most fundamentally, the gap between our canonicalization levels
and full definitional equality.  We hope this work demonstrates that
content-addressing is a productive lens through which to view the
organization of formal mathematical knowledge, and that the tools
presented here are a useful step toward a more interconnected ecosystem
for formalized mathematics.


% =====================================================================
% References
% =====================================================================

\begin{thebibliography}{99}

\bibitem{git}
  L.~Torvalds and J.~Hamano,
  ``Git: Fast version control system,''
  \url{https://git-scm.com}, 2005.

\bibitem{dolstra2004nix}
  E.~Dolstra, M.~de~Jonge, and E.~Visser,
  ``Nix: A safe and policy-free system for software deployment,''
  in \emph{Proc.\ LISA}, USENIX, 2004, pp.~79--92.

\bibitem{benet2014ipfs}
  J.~Benet,
  ``IPFS---Content addressed, versioned, P2P file system,''
  \emph{arXiv preprint arXiv:1407.3561}, 2014.

\bibitem{unison}
  P.~Chiusano and R.~Bjarnason,
  ``Unison: A friendly programming language from the future,''
  \url{https://www.unison-lang.org}, 2019.

\bibitem{lean4}
  L.~de~Moura and S.~Ullrich,
  ``The Lean~4 theorem prover and programming language,''
  in \emph{Proc.\ CADE-28}, Springer LNCS, vol.~12699, 2021,
  pp.~625--635.

\bibitem{coq}
  The Coq Development Team,
  ``The Coq proof assistant,''
  \url{https://coq.inria.fr}, 2024.

\bibitem{isabelle}
  T.~Nipkow, L.~C.~Paulson, and M.~Wenzel,
  \emph{Isabelle/HOL---A Proof Assistant for Higher-Order Logic},
  Springer LNCS, vol.~2283, 2002.

\bibitem{agda}
  U.~Norell,
  ``Towards a practical programming language based on dependent type theory,''
  Ph.D.\ thesis, Chalmers University of Technology, 2007.

\bibitem{mathlib4}
  The Mathlib Community,
  ``Mathlib4: The math library of Lean~4,''
  \url{https://github.com/leanprover-community/mathlib4}, 2024.

\bibitem{ringer2021repair}
  T.~Ringer, R.~Palmkvist, I.~Sergey, M.~Gligoric, and Z.~Tatlock,
  ``Proof repair across type equivalences,''
  in \emph{Proc.\ PLDI}, ACM, 2021, pp.~112--127.

\bibitem{gopinathan2023repair}
  K.~Gopinathan, M.~Sergey, and I.~Sergey,
  ``Mostly automated proof repair for verified libraries,''
  in \emph{Proc.\ PLDI}, ACM, 2023.

\bibitem{coquand_paulin1990}
  T.~Coquand and C.~Paulin-Mohring,
  ``Inductively defined types,''
  in \emph{COLOG-88}, Springer LNCS, vol.~417, 1990, pp.~50--66.

\bibitem{carneiro2019lean}
  M.~Carneiro,
  ``Lean's type theory,''
  \url{https://github.com/digama0/lean-type-theory}, 2019.

\bibitem{debruijn1972}
  N.~G.~de~Bruijn,
  ``Lambda calculus notation with nameless dummies, a tool for
  automatic formula manipulation, with application to the
  Church-Rosser theorem,''
  \emph{Indagationes Mathematicae}, vol.~34, pp.~381--392, 1972.

\bibitem{merkle1988}
  R.~C.~Merkle,
  ``A digital signature based on a conventional encryption function,''
  in \emph{Advances in Cryptology---CRYPTO~'87}, Springer LNCS,
  vol.~293, 1988, pp.~369--378.

\bibitem{kahn1962topological}
  A.~B.~Kahn,
  ``Topological sorting of large networks,''
  \emph{Communications of the ACM}, vol.~5, no.~11, pp.~558--562, 1962.

\bibitem{abel2018decidability}
  A.~Abel, J.~\"{O}hman, and A.~Vezzosi,
  ``Decidability of conversion for type theory in type theory,''
  \emph{Proc.\ ACM Program.\ Lang.}, vol.~2, no.~POPL, 2018,
  pp.~23:1--23:29.

\bibitem{coquand1991algorithm}
  T.~Coquand,
  ``An algorithm for testing conversion in type theory,''
  in \emph{Logical Frameworks}, Cambridge Univ.\ Press, 1991,
  pp.~255--279.

\bibitem{hurd2011opentheory}
  J.~Hurd,
  ``The OpenTheory standard theory library,''
  in \emph{Proc.\ NFM}, Springer LNCS, vol.~6617, 2011, pp.~177--191.

\bibitem{gauthier2019aligning}
  T.~Gauthier, C.~Kaliszyk, and J.~Urban,
  ``Aligning concepts across proof assistant libraries,''
  \emph{J.\ Symbolic Comput.}, vol.~90, pp.~89--103, 2019.

\bibitem{muller2017alignment}
  D.~M\"{u}ller, C.~Gauthier, M.~Kohlhase, and F.~Rabe,
  ``Alignment-based translations across formal systems,''
  2017, manuscript.

\bibitem{rabe2013scalable}
  F.~Rabe and M.~Kohlhase,
  ``A scalable module system,''
  \emph{Information and Computation}, vol.~230, pp.~1--54, 2013.

\bibitem{kohlhase2006omdoc}
  M.~Kohlhase,
  \emph{OMDoc---An Open Markup Format for Mathematical Documents
  (Version~1.2)},
  Springer LNAI, vol.~4180, 2006.

\bibitem{kohlhase2017qed}
  M.~Kohlhase and F.~Rabe,
  ``QED reloaded: Towards a pluralistic formal library of mathematical
  knowledge,''
  \emph{J.\ Formalized Reasoning}, vol.~9, no.~1, pp.~201--234, 2016.

\bibitem{openmath}
  The OpenMath Society,
  ``OpenMath: An extensible standard for representing the semantics of
  mathematical objects,''
  \url{https://openmath.org}, 2004.

\bibitem{assaf2016dedukti}
  A.~Assaf, G.~Burel, R.~Cauderlier, D.~Delahaye, G.~Dowek,
  C.~Dubois, F.~Gilbert, P.~Halmagrand, O.~Hermant, and R.~Saillard,
  ``Dedukti: A logical framework based on the $\lambda\Pi$-calculus
  modulo theory,'' manuscript, 2016.

\bibitem{syntheticthm}
  S.~Agarwal, A.~Welleck, et~al.,
  ``Synthetic theorem generation in Lean,''
  \emph{OpenReview preprint}, 2024.

\bibitem{abel2012long}
  A.~Abel and G.~Scherer,
  ``On irrelevance and algorithmic equality in predicative type theory,''
  \emph{Logical Methods in Computer Science}, vol.~8, no.~1, 2012.

\bibitem{batteries}
  The Lean Community,
  ``Batteries: Community-maintained libraries for Lean~4,''
  \url{https://github.com/leanprover-community/batteries}, 2024.

\bibitem{isabelle_afp}
  G.~Klein, T.~Nipkow, and L.~Paulson, eds.,
  ``Archive of formal proofs,''
  \url{https://www.isa-afp.org}, 2004.

\end{thebibliography}


% =====================================================================
\appendix
\section{Engineering Obstacles}
\label{app:engineering}
% =====================================================================

A practical obstacle to Level~1 canonicalization at scale: Lean's native
\texttt{whnf} (implemented in C) contains no heartbeat or memory
checks.  When processing large environments (${\sim}58{,}000$
declarations in Mathlib modules), two failure modes arise.  First,
\emph{MetaM state accumulation}: a single \texttt{MetaM} session
crashes after ${\sim}25{,}000$ \texttt{whnf} calls due to cumulative
growth of caches, discriminant trees, and instance tables; the crash
point depends on call count, not on which specific declarations are
processed.  Second, \emph{process-level memory growth}: applying
\texttt{whnf} to definition values causes monotonic RSS growth that
eventually crashes regardless of batch boundaries.

\paragraph{Micro-batching.}  Our workaround is to process declarations
in micro-batches of~10, each in a fresh \texttt{MetaM} session.  Types
receive Level~1 normalization; values are always canonicalized with
Level~0 only (no \texttt{whnf}).  For theorems, the value hash is the
sentinel \texttt{PROOF\_IRRELEVANT} (proof irrelevance).  The result is
an asymmetric strategy: \textbf{types get L1, values get L0}.  Hash
computation is split into two phases (canonicalize+hash vs.\
pretty-print) in separate \texttt{MetaM} sessions, with a flush every
5{,}000 declarations to bound live \texttt{Expr} objects in memory.

An upstream fix---heartbeat checks and resource guards in Lean's native
\texttt{whnf} implementation---would eliminate the need for
micro-batching entirely.


\end{document}
